---
title: "Meen & Whitehead 2020"
#author: "Jens von Bergmann"
date: "13/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width=9)
library(tidyverse)
library(patchwork)
```


## Housing expenditures to income

Figure 2.1 on page 21 looks at housing expenditure to income ratios in England. We reproduce this for Metro Vancouver using 2016 (individuals) PUMF data, utilizing the shelter cost, household income, after-tax household income and government transfers variables.


```{r}
# path to 2016 pumf data (spss format)
path=file.path(getOption("custom_data_path"),"pumf/pumf_2016/Data","Census_2016_Individual_PUMF.sav") 
pumf_data <- foreign::read.spss(path) %>%
  as_tibble() %>%
  set_names(names(.) %>% toupper) %>%
  mutate(housing=SHELCO*12) %>%
  mutate(lower=case_when(HHINC=="Under $2,000" ~ 0,
                         TRUE ~ as.numeric(gsub("\\$|,","",str_extract(as.character(HHINC),"^\\$[\\d,]+")))),
         upper=case_when(HHINC=="$250,000 and over" ~ 350000,
                         TRUE~as.numeric(gsub("\\$|,","",str_extract(as.character(HHINC),"\\$[\\d,]+$"))))) %>%
  mutate(income=(lower+upper)/2) %>%
  mutate(lower2=case_when(HHINC_AT=="Under $2,000" ~ 0,
                         TRUE ~ as.numeric(gsub("\\$|,","",str_extract(as.character(HHINC_AT),"^\\$[\\d,]+")))),
         upper2=case_when(HHINC_AT=="$250,000 and over" ~ 300000,
                         TRUE~as.numeric(gsub("\\$|,","",str_extract(as.character(HHINC_AT),"\\$[\\d,]+$"))))) %>%
  mutate(income_at=(lower2+upper2)/2+GTRFS) %>%
  filter(!is.na(income),!is.na(income_at),!is.na(housing),TENUR!="Not available") %>%
  mutate(scir=pmin(1,housing/income),
         scir_at=pmin(1,housing/income_at))

plot_theme <- list(
  geom_boxplot(outlier.shape=NA),
  theme_light(),
  scale_fill_manual(values=c("Owned by a member of the household"="steelblue","Rented or Band housing"="brown"),
                    guide=FALSE),
  facet_wrap("TENUR",ncol=1),
  theme(axis.text.x = element_text(angle = 90,hjust=1))
)

housing_income_plot <- function(d){
  p1 <- ggplot(d,aes(x=HHINC,y=scir,fill=TENUR)) +
    plot_theme +
    labs(title="Total income",
         x=NULL,y="Shelter cost to income ratio")
  
  p2 <- ggplot(d,aes(x=HHINC_AT,y=scir_at,fill=TENUR)) +
    plot_theme +
    labs(title="After-tax income, including transfers",
         x=NULL,y=NULL)
  
  p1 + p2
}

```

```{r}
pumf_data %>%
  filter(CMA=="Vancouver",
         PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() + plot_annotation(
  title = 'Metro Vancouver housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

We see some effect of moderation in shelter cost to income ratios via government transfers for lower income renter households.

National level data gives a cleaner comparison to the graph in Meek and Whitehead, and the effect is a little more pronounced.

```{r}
pumf_data %>%
  filter(PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() +  plot_annotation(
  title = 'Canada housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

For comparison we also look at Toronto and Calgary.


```{r}
pumf_data %>%
  filter(CMA=="Toronto",
         PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() +  plot_annotation(
  title = 'Toronto housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

```{r}
pumf_data %>%
  filter(CMA=="Calgary",
         PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() +  plot_annotation(
  title = 'Calgary housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

### Movers 5
Another way to look at this is to only consider people that moved recently (in the past 5 years), excluding external migrants because it usually takes a couple of years until their income stabelizes.

```{r}
pumf_data %>%
  filter(CMA=="Vancouver",
         PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         MOB5 %in% c("Non-migrants", "Different CSD, same census division",
                     "Different CD, same province", "Interprovincial migrants"),
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() + plot_annotation(
  title = 'Metro Vancouver internal movers 5 housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

```{r}
pumf_data %>%
  filter(PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         MOB5 %in% c("Non-migrants", "Different CSD, same census division",
                     "Different CD, same province", "Interprovincial migrants"),
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() + plot_annotation(
  title = 'Canada internal movers 5 housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```
