---
title: "Meen & Whitehead 2020"
#author: "Jens von Bergmann"
date: "13/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width=9)
library(tidyverse)
library(patchwork)
```


## Housing expenditures to income

Figure 2.1 on page 21 looks at housing expenditure to income ratios in England. We reproduce this for Metro Vancouver using 2016 (individuals) PUMF data, utilizing the shelter cost, household income, after-tax household income and government transfers variables.


```{r}
# path to 2016 pumf data (spss format)
path=file.path(getOption("custom_data_path"),"pumf/pumf_2016/Data","Census_2016_Individual_PUMF.sav") 
pumf_data <- foreign::read.spss(path) %>%
  as_tibble() %>%
  set_names(names(.) %>% toupper) %>%
  mutate(housing=SHELCO*12) %>%
  mutate(lower=case_when(HHINC=="Under $2,000" ~ 0,
                         TRUE ~ as.numeric(gsub("\\$|,","",str_extract(as.character(HHINC),"^\\$[\\d,]+")))),
         upper=case_when(HHINC=="$250,000 and over" ~ 350000,
                         TRUE~as.numeric(gsub("\\$|,","",str_extract(as.character(HHINC),"\\$[\\d,]+$"))))) %>%
  mutate(income=(lower+upper)/2) %>%
  mutate(lower2=case_when(HHINC_AT=="Under $2,000" ~ 0,
                         TRUE ~ as.numeric(gsub("\\$|,","",str_extract(as.character(HHINC_AT),"^\\$[\\d,]+")))),
         upper2=case_when(HHINC_AT=="$250,000 and over" ~ 300000,
                         TRUE~as.numeric(gsub("\\$|,","",str_extract(as.character(HHINC_AT),"\\$[\\d,]+$"))))) %>%
  mutate(income_at=(lower2+upper2)/2+GTRFS) %>%
  filter(!is.na(income),!is.na(income_at),!is.na(housing),TENUR!="Not available") %>%
  mutate(scir=pmin(1,housing/income),
         scir_at=pmin(1,housing/income_at))

plot_theme <- list(
  geom_boxplot(outlier.shape=NA),
  theme_light(),
  scale_fill_manual(values=c("Owned by a member of the household"="steelblue","Rented or Band housing"="brown"),
                    guide=FALSE),
  facet_wrap("TENUR",ncol=1),
  theme(axis.text.x = element_text(angle = 90,hjust=1))
)

housing_income_plot <- function(d){
  p1 <- ggplot(d,aes(x=HHINC,y=scir,fill=TENUR)) +
    plot_theme +
    labs(title="Total income",
         x=NULL,y="Shelter cost to income ratio")
  
  p2 <- ggplot(d,aes(x=HHINC_AT,y=scir_at,fill=TENUR)) +
    plot_theme +
    labs(title="After-tax income, including transfers",
         x=NULL,y=NULL)
  
  p1 + p2
}

```

```{r}
pumf_data %>%
  filter(CMA=="Vancouver",
         PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() + plot_annotation(
  title = 'Metro Vancouver housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

We see some effect of moderation in shelter cost to income ratios via government transfers for lower income renter households.

National level data gives a cleaner comparison to the graph in Meek and Whitehead, and the effect is a little more pronounced.

```{r}
pumf_data %>%
  filter(PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() +  plot_annotation(
  title = 'Canada housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

For comparison we also look at Toronto and Calgary.


```{r}
pumf_data %>%
  filter(CMA=="Toronto",
         PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() +  plot_annotation(
  title = 'Toronto housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

```{r}
pumf_data %>%
  filter(CMA=="Calgary",
         PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() +  plot_annotation(
  title = 'Calgary housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

### Movers 5
Another way to look at this is to only consider people that moved recently (in the past 5 years), excluding external migrants because it usually takes a couple of years until their income stabelizes.

```{r}
pumf_data %>%
  filter(CMA=="Vancouver",
         PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         MOB5 %in% c("Non-migrants", "Different CSD, same census division",
                     "Different CD, same province", "Interprovincial migrants"),
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() + plot_annotation(
  title = 'Metro Vancouver internal movers 5 housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

```{r}
pumf_data %>%
  filter(PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         MOB5 %in% c("Non-migrants", "Different CSD, same census division",
                     "Different CD, same province", "Interprovincial migrants"),
         #!(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         SHELCO>0) %>% # don't consider households without shelter cost
  housing_income_plot() + plot_annotation(
  title = 'Canada internal movers 5 housing expenditure to household income ratios',
  subtitle = '(non-student lead households with shelter cost greater than 0)',
  caption = 'StatCan 2016 Census Individuals PUMF'
)
```

```{r}
cmas <- as.vector(unique(pumf_data$CMA)) # Extract each CMA

lorenz_curve <- function(`CMA`) {  
## Get income distribution for potential first-time buyers
pumf_data_renters_u50 <- pumf_data %>%
  filter(PRIHM=="Person is primary maintainer", # count households only once
         ATTSCH=="Did not attend school", # filter out students
         GTRFS!=88888888,
         GTRFS!=99999999,
         !(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")),
         !(MOB1 %in% c("External migrants", "Not available")),  #Remove 1st year external migrants
         SHELCO>0, 
         !(AGEGRP %in% c("50 to 54 years", "55 to 59 years" ,"60 to 64 years", "65 to 69 years", "70 to 74 years","75 to 79 years", "80 to 84 years", "85 years and over", "Not available")), 
         TENUR=="Rented or Band housing",
         CMA==`CMA`)

## Get price distribution for properties reasonably affordable to first-time buyers
pumf_data_owners <- pumf_data %>%
  filter(PRIHM=="Person is primary maintainer", # count households only once
         !(SUBSIDY %in% c("Yes, a subsidized dwelling", "Not available")), #should be redundant when looking at owners
         SHELCO>0, 
         TENUR=="Owned by a member of the household",
         DTYPE!= "Not available",
         CMA==`CMA`) %>%
    mutate(`Annual mortgage cost`= case_when(VALUE < 1000000 ~ 12*pmt(r=(0.0225/12),pv=-0.99*VALUE, n=300, fv=0), #.99 is 5% down with 4% mortgage insurance  financed
                                             VALUE >=1000000 ~ 12*pmt(r=(0.025/12),pv=-0.8*VALUE, n=360, fv=0))) %>%
    filter(`Annual mortgage cost`< 0.3*max(pumf_data_renters_u50$income))  

## Create dataframe for Lorenz curve
ownership_affordability_`CMA` <- as.tibble(quantile(pumf_data_renters_u50$income, seq(0,1,0.1))) %>%
  transmute(Income=value) %>%
  mutate(`Max affordable housing cost`=0.3*Income, 
          `Can affordably buy up to X price percentile`= ecdf(pumf_data_owners$`Annual mortgage cost`)(`Max affordable housing cost`),
         Price=quantile(pumf_data_owners$VALUE, `Can affordably buy up to X price percentile`))
ownership_affordability$`Income percentile` = seq(0,1,0.1)
ownership_affordability$CMA = match.call()

}
        
ggplot(ownership_affordability, aes(x=`Income percentile`, y=`Can affordably buy up to X price percentile`)) +
  geom_point() +
  geom_abline(slope = 1) +
  geom_smooth(se=FALSE, span=0.5) +
  theme_minimal() + 
  labs(title = "Vancouver CMAs first-time home buyer Lorenz curve",
       caption = "Statistics Canada 2016 Individual Census PUMF \n First-time home buyer income is proxied by using household income for under 50 years old non-subsidized, non-student, non-first year external migr renter households \n Affordability set at 30% of household income \n Assumes 5% down payment, 25 year amortization, and 2.25% mortgage rate for sub $1 million properties and \n 20% down payment, 30 year amortization, and 2.5% mortgage rate for  $1 million + properties \n Distribution of prices truncated to affordable to max of income for under 60 years old non-subsidized renter households") +
  xlab("Income percentile of potential first-time home buyers") +
  ylab("Can affordably buy up to X of truncated price percentile")
```


